\textsf{Are large language models superhuman chemists?}

### 2024-03-27 src/tex/ms.tex
reading pass

    We curated more than 7000 question answer-pairs about a wide array of subfields, evaluated leading models and found that the best models, on average, outperform the best human chemists in our study. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and provide overconfident misleading predictions, e.g., about the safety profile of chemicals.

    These findings underscore the dual reality that, although LLMs demonstrate remarkable proficiency in chemical tasks, there is a critical need for further research to enhance their safety and utility in chemical sciences.
    Our findings also indicate a need for adaptation to chemistry curricula and highlight the importance of continued development of evaluation frameworks for systematic improvement of safe and useful LLMs.
\end{abstract}

\clearpage
\begin{refsection}
\section{Introduction}
\Glspl{llm} are \gls{ml} models trained on massive amounts of text to complete sentences. 
Aggressive scaling of these models has led to a rapid increase in their capabilities,\autocite{brown2020language,zhong2024benchmarking} with the leading models now being able to pass in some evaluations the United States Medical Licensing Examination\autocite{kung2023performance} or identify and autonomously buy chemicals when augmented with external tools such as web search and synthesis planners.\autocite{openai2024gpt4}
While some see \enquote{sparks of \gls{agi}} in them,\autocite{bubeck2023sparks} others consider them as \enquote{stochastic parrots}---i.e., systems that only regurgitate what they have been trained on.\autocite{bender2021dangers}
In any case, the promise of so-called foundation models (the leading ones being called frontier models) is that they show the ability to solve a wide variety of tasks they have not been explicitly trained on.\autocite{bommasani2021opportunities, anderljung2023frontier}
This led to tremendous economic interest and investment in such generative models, with an expected market of more than \$1.3 trillion (almost \$30 billion for drug discovery applications) by 2032.\autocite{bloomberg}

Chemists and materials scientists have quickly caught on the mounting attention given to \glspl{llm}, with some voices even suggesting that \enquote{the future of chemistry is language}.\autocite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to predict properties of molecules or materials,\autocite{jablonka202314, jablonka2024leveraging, xie2024fine, liao2024words, zhang2024chemllm, zhong2024benchmarking}  optimize reactions\autocite{ramos2023bayesian, kristiadi2024sober}  generate materials,\autocite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\autocite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured} or to even build first prototypes of systems that can autonomously perform reactions in the physical world based on commands provided in natural language.\autocite{bran2023chemcrow, Boiko_2023, darvish2024organa}

In addition, since a lot---if not most---of the information about chemistry is currently stored and communicated in text there is a strong reason to believe that there is a lot of untapped potential in \glspl{llm} for chemistry and materials science.
For instance, most insights in chemical research do not directly originate from data stored in databases but rather from the scientists and their ability to interpret data. 
Many of these insights are in form of text in scientific publications. 
Thus, operating on such texts might be our best way of unlocking these insights and to learn from them.
This might ultimately lead to general copilot systems for chemists that can provide answers to questions or even suggest new experiments based on vastly more information than a human could ever read.
Such a usage mode is in particular interesting in the face of recent advances in autonomous laboratories.\autocite{Boiko_2023, bran2023chemcrow, darvish2024organa, granda2018controlling, Angello_2022, coley2019robotic, Burger_2020, seifrid2022autonomous}

However, the rapid increase in capabilities of chemical \gls{ml} models led (even before the recent interest in \glspl{llm}) to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\autocite{gopal2023releasing, ganguli2022red, Urbina_2022, campbell2023censoring, moulange2023towards, urbina2022teachable}
To some extent, this is not surprising as any technology that, for instance, is used to design non-toxic molecules can also be used in the inverse way to predict toxic ones. Yet, it is important to realize that, while \glspl{llm} can facilitate access to information and tools, this information about controlled chemicals and task-specific tools have been available for years.
However, it is important to realize that the user base of such models is not only limited to experts in chemistry and materials science who critically reflect on every output such models produce. 
For example, many students frequently consult these tools---perhaps even to prepare chemical experiments.\autocite{Intelligent.com_2023}
This also applies to users from the general public, who might consider using \glspl{llm} to answer questions about the safety of chemicals.
Thus, for some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. 
However, even for experts the chemical understanding and reasoning capabilities are important as they will determine the capabilities and limitations of their models in their work, e.g., in copilot systems for chemists.
Unfortunately, apart from anecdotal reports, there is little evidence on how \glspl{llm} perform compared to experts.

Thus, to better understand what \glspl{llm} can do for chemistry and materials science, and where they might be improved with further developments, evaluation frameworks are needed to allow us to systematically measure progress and mitigate potential harms.
For the development of \glspl{llm}, evaluation is currently mostly performed via standardized benchmarks suites such as BigBench\autocite{srivastava2022beyond} or the LM Eval Harness.\autocite{eval-harness}
The former contains, among 204 tasks (such as linguistic puzzles), only two tasks classified as \enquote{chemistry related}, whereas the latter contains no specific chemistry tasks.
Due to the lack of widely accepted standard benchmarks, the developers of chemical language models\autocite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024, frey2023neural} frequently utilize language-interfaced\autocite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\autocite{wu2018moleculenet} Therapeutic Data Commons\autocite{huang2021therapeutics} or MatBench.\autocite{dunn2020benchmarking}
In these cases, the models are  evaluated on predicting very specific properties of molecules (e.g., solubility, toxicity, melting temperature or reactivity) or on predicting the outcome of specific chemical reactions.
This, however, only gives a very limited view of the general chemical capabilities of the models.

@@

\section{Results}

@@

To compile our benchmark corpus we utilized a broad list of sources (see \Cref{sec:curation}), ranging from university exams to semi-automatically generated questions based on curated subsets of data in chemical databases.
For quality assurance, all questions have been reviewed by at least one scientist in addition to the original curator as well as automatic checks.

@@

\paragraph{\enquote{Tiny} subset}
It is important to note that for routine evaluations, a smaller subset of the corpus might be more practical.\autocite{polo2024tinybenchmarks}
For instance,~\textcite{liang2023holistic} report costs of more than 10,000 USD for \gls{api} calls for a single evaluation on the widely used \gls{helm} benchmark. 
To address this, we also provide a subset (\variable{output/num_tiny_questions.txt} questions) of the corpus that was curated to be a diverse and representative subset of the full corpus in which topics are more balanced than in the complete corpus (see \Cref{sec:subset-selection} for details on the curation process).
We also used this subset to seed the web application for the human baseline study.

@@

\paragraph{Benchmark suite design} Because the text used in scientific settings is different from typical natural language, many models have been developed that deal with such text in a special way.
For instance, Galactica\autocite{taylor2022galactica} uses special tokenization or encoding procedures for molecules and equations. 
Current benchmarking suites, however, do not account for such special treatment of scientific information.
To address this, \chembench is designed to encode the semantic meaning of various parts of the question or answer.  
For instance, molecules represented in \gls{smiles} are enclosed in \texttt{[START\_SMILES][\textbackslash END\_SMILES]} tags. 
This allows the model to treat the \gls{smiles} string different from other text. 
\chembench can seamlessly handle such special treatment in an easily extensible way because the questions are stored in an annotated format.

@@

Given the large interest in tool-augmented systems, the bad performance of these systems (GPT-3.5 and Claude 2 augmented with tools) in our benchmark is striking. 
Their lack of performance, however, is partially due to the fact that we limited the system to a maximum of ten \gls{llm} calls.
With the default tool augmented setup (using the so-called ReAct method\autocite{yao2023react}), however, this often did not allow the system to identify the correct answer (e.g., because it repeatedly tried to search for the answer on the web and did not find a solution within ten calls to the \gls{llm}).
This observation highlights the importance of communicating and measuring not only in predictive performance but also computational cost (e.g., in terms of \gls{api} calls) for tool-augmented systems.

% Radar chart
\paragraph{Performance per topic} To obtain a more detailed understanding of the performance of the models, we also analyzed the performance of the models in different subfields of the chemical sciences.
For this analysis, we defined a set of topics (see \Cref{sec:meth-topic}) and classified all questions in the \chembench corpus into these topics based on hand-crafted rules as well as classifier models operating on the question texts.
We then computed the percentage of questions that the models, or humans, answered completely correctly for each topic.
\begin{figure}[!h]
    \centering
    \includegraphics{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics of the \enquote{tiny} subset of the \chembench corpus.} The radar plot shows the performance of the models and humans on the different topics of \enquote{tiny} subset of the \chembench corpus. The performance is measured as the fraction of questions that were answered completely correct by the models.
    The best score for every dimension is one (all questions answered correctly) and the worst score is zero (no question answered completely correctly). A larger colored area indicates a better performance.
    This figure shows the performance on the subset of questions that was answered by humans. The performance of models on the entire corpus is shown in \Cref{fig:performance_per_topic}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}
In this spider chart, the worst score for every dimension is zero (no question answered completely correctly) and the best score is one (all questions answered correctly). 
Thus, a larger colored area indicates a better performance. 
One can observe that this performance varies widely across models and topics. 
While macromolecular chemistry and biochemistry receive relatively high scores for many models, this is not the case for topics such as chemical safety or analytical chemistry.
In the subfield of analytical chemistry the prediction of the number of signals observable in a \gls{nmr} spectrum proved difficult for the models (e.g., \variable{output/subset_scores/is_number_nmr_peaks_gpt4.txt} percent correct answers for GPT-4) while this question appeared easier (\variable{output/human_subset_scores/is_number_nmr_peaks.txt} percent correct) for trained humans.
Importantly, the human experts are given a drawing of the compounds, whereas models are only shown the \gls{smiles} string of a compound and have to use this to reason about the symmetry of the compound (i.e., to identify the number of diasterotopically distinct protons, which requires \emph{reasoning} about the topology and structure of a molecule). 
These findings also shine an interesting light on the value of textbook-inspired questions. 
A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students. 
On those questions the models tend to perform better than on some of our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
For instance, while the overall performance in the chemical safety topic is low, the models would pass the certification exam according to the German Chemical Prohibition Ordinance based on a subset of a questions we sampled from the corresponding question bank (e.g., \variable{output/subset_scores/is_gfk_gpt4.txt}\% correct answers for GPT-4, \variable{output/subset_scores/is_gfk_claude3.txt}\% for Claude 3, and \variable{output/human_subset_scores/is_gfk.txt}\% for the human experts).
While those findings are impacted by the subset of questions we sampled, the results still highlight that good performance on such question bank or textbook questions does not necessarily translate to good performance on other questions that require more reasoning.

@@

\begin{figure}[!h]
    \centering
    \includegraphics{figures/confidence_vs_performance_overall.pdf}
    \caption{\textbf{Relationship between confidence of the model in the answer and correctness.} For this analysis, we used verbalized confidence estimates from the model. To obtain those estimates, we prompted the models to return a confidence score on an ordinal scale. We then plot the correctness of the answers (which is calculated as the mean value of all answers being either completely correct (1) or completely incorrect (0)) against the confidence score.
        The stripplots show the individual data points, a strong density indicating a higher number of points. The lines show the mean and the errorbars show the standard error of the mean. The figure shows that the reliability of the confidence estimates varies widely across models. While the ones for GPT-4 and Claude-2 are not particularly reliable, the ones for Claude-3 follow the expected trend.}
    \label{fig:confidence_vs_performance}
    \script{joint_analysis_confidence_performance.py}
\end{figure}

@@

\section{Discussion and Conclusions}
On one hand, our findings show that it is impressive how far we got. Leading models outperform domain experts in chemistry questions on many topics. 
On the other hand, there are still striking limitations. On very relevant topics the answers models provide are wrong. And models are not able to estimate their own limitations.
However, the success of our benchmark perhaps also reveals more about the limitations of the exams we use to evaluate models---and chemists---than about the models themselves.
For instance, while models perform well on many textbook questions, they struggle with questions that require some more reasoning.

Yet, given that the models outperform the average human in our study, it is clear that we need to rethink the way we teach and examine chemistry.
Critical reasoning is increasingly important and rote solving of problems or memorization of facts is clearly a domain in which \glspl{llm} will continue to outperform humans.

Our findings also highlight the nuanced trade-off between breadth and depth of evaluation frameworks. 
The analysis of model performance on different topics shows that the performance of models varies widely across topics. 
But even within a topic, the performance of models can vary widely depending on the type of question and the reasoning required to answer it.

The current evaluation frameworks for \glspl{llm} are mostly designed to measure the performance of the models on specific tasks. 
They cannot easily be used to evaluate systems, or models built for scientific applications. 
Thus, we had only little understanding of the capabilities of \glspl{llm} in the chemical sciences.
Our work, shows that carefully curated benchmarks can provide a more nuanced understanding of the capabilities of \glspl{llm} in the chemical sciences.
It lays the groundwork for the development of more advanced systems that can be used in the chemical sciences and highlights the importance of continued development of evaluation frameworks for systematic improvement of safe and useful \glspl{llm}.
Our findings also illustrate that more focus is required in the development of better human-model interaction frameworks given that models are not able to estimate their own limitations.

While our findings indicate many areas for further improvement of \gls{llm}-based systems, it is also important to realize that clearly defined metrics have been the key for the progress of many fields of \gls{ml} such computer vision. 
We thus believe that although current systems are far from reasoning like a chemist our open source \chembench framework will be a stepping stone for the development of systems that might come closer to this goal.

@@

\subsection{Curation workflow}\label{sec:curation}
For our dataset we curated questions from existing exams, but also programmatically created new questions.
Questions were added via Pull Requests on our GitHub repository and only merged into the corpus after passing manual review.

@@

\paragraph{Parsing}
Our parsing workflow is multistep and primarily based on regular expressions.
In the case of instruction-tuned models, we first attempt to identify the \texttt{[ANSWER][\textbackslash ANSWER]} environment we prompt the model to report the answer in.
In the case of completion models, this step is skipped. From there, we attempt to extract the relevant enumeration letters (for multiple-choice questions) or numbers.
In the case of numbers, our regular expression was engineered to be able to deal with various forms of scientific notation.
As initial tests indicated that models sometimes return integers in the form of words, e.g. \enquote{one} instead of \enquote{1}, we also implemented a word-to-number conversion using regular expressions.
In case these hard-coded parsing steps fail, we fall back to using a \gls{llm}, e.g. Claude-2, to parse the completion.
%Manual verification indicates that this \gls{llm}-based parsing represent not a relevant error source and has been performed correctly in all cases we checked (randomly sampled 4 questions per topic).

@@

M.A. expresses gratitude to the European Research Council (ERC) for evaluating the project with the reference number 101106377 titled \enquote{CLARIFIER} and accepting it for funding under the HORIZON TMA MSCA Postdoctoral Fellowships - European Fellowships. 
Furthermore, M.A. acknowledges the funding provided by UK Research and Innovation (UKRI) under the UK government’s Horizon Europe funding guarantee (Grant Reference: EP/Y023447/1; Organization Reference: 101106377).

M.R. and U.S.S. thanks the \enquote{Deutsche Forschungsgemeinschaft} for funding under the regime of the priority programme SPP 2363 \enquote{Utilization and Development of Machine Learning for Molecular Applications – Molecular Machine Learning} (SCHU 1229/63-1; project number 497115849).

>>>>

    We curated more than 7000 question-answer pairs about a wide array of subfields of the chemical sciences, evaluated leading models and found that the best models, on average, outperform the best human chemists in our study. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and provide overconfident misleading predictions, e.g., about the safety profile of chemicals.

    These findings underscore the dual reality that, although LLMs demonstrate remarkable proficiency in chemical tasks, there is a critical need for further research to enhance their safety and utility in chemical sciences.
    Our findings also indicate a need for adaptations to chemistry curricula and highlight the importance of continued development of evaluation frameworks for systematic improvement of safe and useful LLMs.
\end{abstract}

\clearpage
\begin{refsection}
\section{Introduction}
\Glspl{llm} are \gls{ml} models trained on massive amounts of text to complete sentences. 
Aggressive scaling of these models has led to a rapid increase in their capabilities,\autocite{brown2020language,zhong2024benchmarking} with the leading models now being able to pass in some evaluations the United States Medical Licensing Examination.\autocite{kung2023performance} 
They also have been shown to design and autonomously perform chemical reactions when augmented with external tools such as web search and synthesis planners.\autocite{openai2024gpt4, Boiko_2023, bran2023chemcrow}
While some see \enquote{sparks of \gls{agi}} in them,\autocite{bubeck2023sparks} others consider them as \enquote{stochastic parrots}---i.e., systems that only regurgitate what they have been trained on.\autocite{bender2021dangers}
In any case, the promise of these models is that they show the ability to solve a wide variety of tasks they have not been explicitly trained on.\autocite{bommasani2021opportunities, anderljung2023frontier}
This led to tremendous economic interest and investment in such generative models, with an expected market of more than \$1.3 trillion (almost \$30 billion for drug discovery applications) by 2032.\autocite{bloomberg}

Chemists and materials scientists have quickly caught on the mounting attention given to \glspl{llm}, with some voices even suggesting that \enquote{the future of chemistry is language}.\autocite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to predict properties of molecules or materials,\autocite{jablonka202314, jablonka2024leveraging, xie2024fine, liao2024words, zhang2024chemllm, zhong2024benchmarking}  optimize reactions\autocite{ramos2023bayesian, kristiadi2024sober}  generate materials,\autocite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\autocite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured} or to even prototype systems that can autonomously perform reactions in the physical world based on commands provided in natural language.\autocite{bran2023chemcrow, Boiko_2023, darvish2024organa}

In addition, since a lot---if not most---of the information about chemistry is currently stored and communicated in text there is a strong reason to believe that there is still a lot of untapped potential in \glspl{llm} for chemistry and materials science.
For instance, most insights in chemical research do not directly originate from data stored in databases but rather from the scientists and their ability to interpret data. 
Many of these insights are in form of text in scientific publications. 
Thus, operating on such texts might be our best way of unlocking these insights and to learn from them.
This might ultimately lead to general copilot systems for chemists that can provide answers to questions or even suggest new experiments based on vastly more information than a human could ever read.
Such a usage mode is in particular interesting in the face of recent advances in autonomous laboratories.\autocite{Boiko_2023, bran2023chemcrow, darvish2024organa, granda2018controlling, Angello_2022, coley2019robotic, Burger_2020, seifrid2022autonomous}

However, the rapid increase in capabilities of chemical \gls{ml} models led (even before the recent interest in \glspl{llm}) to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\autocite{gopal2023releasing, ganguli2022red, Urbina_2022, campbell2023censoring, moulange2023towards, urbina2022teachable}
To some extent, this is not surprising as any technology that, for instance, is used to design non-toxic molecules can also be used in the inverse way to predict toxic ones. 
Yet, it is important to realize that, while \glspl{llm} can facilitate access to information and tools, this information about controlled chemicals and task-specific tools have been available for years.
Still, it is important to realize that the user base of such models is not only limited to experts in chemistry and materials science who critically reflect on every output such models produce. 
For example, many students frequently consult these tools---perhaps even to prepare chemical experiments.\autocite{Intelligent.com_2023}
This also applies to users from the general public, who might consider using \glspl{llm} to answer questions about the safety of chemicals.
Thus, for some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. 
However, even for experts the chemical understanding and reasoning capabilities are important as they will determine the capabilities and limitations of their models in their work, e.g., in copilot systems for chemists.
Unfortunately, apart from anecdotal reports, there is little evidence on how \glspl{llm} perform compared to expert (human) chemists.

Thus, to better understand what \glspl{llm} can do for the chemical sciences, and where they might be improved with further developments, evaluation frameworks are needed to allow us to systematically measure progress and mitigate potential harms.
For the development of \glspl{llm}, evaluation is currently mostly performed via standardized benchmark suites such as BigBench\autocite{srivastava2022beyond} or the LM Eval Harness.\autocite{eval-harness}
The former contains, among 204 tasks (such as linguistic puzzles), only two tasks classified as \enquote{chemistry related}, whereas the latter contains no specific chemistry tasks.
Due to the lack of widely accepted standard benchmarks, the developers of chemical language models\autocite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024, frey2023neural} frequently utilize language-interfaced\autocite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\autocite{wu2018moleculenet} Therapeutic Data Commons\autocite{huang2021therapeutics} or MatBench.\autocite{dunn2020benchmarking}
In these cases, the models are  evaluated on predicting very specific properties of molecules (e.g., solubility, toxicity, melting temperature or reactivity) or on predicting the outcome of specific chemical reactions.
This, however, only gives a very limited view of the general chemical capabilities of the models.

@@

\section{Results and Discussion}

@@

To compile our benchmark corpus we utilized a broad list of sources (see \Cref{sec:curation}), ranging from university exams to semi-automatically generated questions based on curated subsets of data in chemical databases.
For quality assurance, all questions have been reviewed by at least one scientist in addition to the original curator as well as automated checks.

@@

\paragraph{\enquote{Tiny} subset}
It is important to note that for routine evaluations, a smaller subset of the corpus might be more practical.\autocite{polo2024tinybenchmarks}
For instance,~\textcite{liang2023holistic} report costs of more than \$10,000 for \gls{api} calls for a single evaluation on the widely used \gls{helm} benchmark. 
To address this, we also provide a subset (\variable{output/num_tiny_questions.txt} questions) of the corpus that was curated to be a diverse and representative subset of the full corpus in which topics are more balanced than in the complete corpus (see \Cref{sec:subset-selection} for details on the curation process).
We also used this subset to seed the web application for the human baseline study.

@@

\paragraph{Benchmark suite design} Because the text used in scientific settings is different from typical natural language, many models have been developed that deal with such text in a special way.
For instance, the Galactica model\autocite{taylor2022galactica} uses special tokenization or encoding procedures for molecules and equations. 
Current benchmarking suites, however, do not account for such special treatment of scientific information.
To address this, \chembench is designed to encode the semantic meaning of various parts of the question or answer.  
For instance, molecules represented in \gls{smiles} are enclosed in \texttt{[START\_SMILES][\textbackslash END\_SMILES]} tags. 
This allows the model to treat the \gls{smiles} string different from other text. 
\chembench can seamlessly handle such special treatment in an easily extensible way because the questions are stored in an annotated format.

@@

Given the large interest in tool-augmented systems, the mediocre performance of these systems (GPT-3.5 and Claude 2 augmented with tools) in our benchmark is striking. 
Their lack of performance, however, is partially due to the fact that we limited the system to a maximum of ten \gls{llm} calls.
With the default tool augmented setup (using the so-called ReAct method\autocite{yao2023react}), however, this often did not allow the system to identify the correct answer (e.g., because it repeatedly tried to search for the answer on the web and did not find a solution within ten calls to the \gls{llm}).
This observation highlights the importance of communicating and measuring not only in predictive performance but also computational cost (e.g., in terms of \gls{api} calls) for tool-augmented systems.

\paragraph{Performance per topic} To obtain a more detailed understanding of the performance of the models, we also analyzed the performance of the models in different subfields of the chemical sciences.
For this analysis, we defined a set of topics (see \Cref{sec:meth-topic}) and classified all questions in the \chembench corpus into these topics based on hand-crafted rules as well as classifier models operating on the question texts.
We then computed the percentage of questions that the models, or humans, answered completely correctly for each topic.
\begin{figure}[!h]
    \centering
    \includegraphics{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics of the \enquote{tiny} subset of the \chembench corpus.} The radar plot shows the performance of the models and humans on the different topics of \enquote{tiny} subset of the \chembench corpus. The performance is measured as the fraction of questions that were answered completely correct by the models.
    The best score for every dimension is one (all questions answered correctly) and the worst score is zero (no question answered completely correctly). A larger colored area indicates a better performance.
    This figure shows the performance on the subset of questions that was answered by humans. The performance of models on the entire corpus is shown in \Cref{fig:performance_per_topic}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}
In this spider chart, the worst score for every dimension is zero (no question answered correctly) and the best score is one (all questions answered correctly). 
Thus, a larger colored area indicates a better performance. 
One can observe that this performance varies widely across models and topics. 
While macromolecular chemistry and biochemistry receive relatively high scores for many models, this is not the case for topics such as chemical safety or analytical chemistry.
In the subfield of analytical chemistry the prediction of the number of signals observable in a \gls{nmr} spectrum proved difficult for the models (e.g., \variable{output/subset_scores/is_number_nmr_peaks_gpt4.txt} percent correct answers for GPT-4) while this question appeared easier (\variable{output/human_subset_scores/is_number_nmr_peaks.txt} percent correct) for trained humans.
Importantly, the human experts are given a drawing of the compounds, whereas models are only shown the \gls{smiles} string of a compound and have to use this to reason about the symmetry of the compound (i.e., to identify the number of diasterotopically distinct protons, which requires \emph{reasoning} about the topology and structure of a molecule). 
These findings also shine an interesting light on the value of textbook-inspired questions. 
A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students. 
On those questions the models tend to perform better than on some of our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
For instance, while the overall performance in the chemical safety topic is low, the models would pass the certification exam according to the German Chemical Prohibition Ordinance based on a subset of questions we sampled from the corresponding question bank (e.g., \variable{output/subset_scores/is_gfk_gpt4.txt}\% correct answers for GPT-4, \variable{output/subset_scores/is_gfk_claude3.txt}\% for Claude 3, and \variable{output/human_subset_scores/is_gfk.txt}\% for the human experts).
While those findings are impacted by the subset of questions we sampled, the results still highlight that good performance on such question bank or textbook questions does not necessarily translate to good performance on other questions that require more reasoning.

We also gain insight in the struggles the models have with chemical reasoning tasks when we look at the performance as a function of molecular descriptors.
If the model would answer questions based on reasoning about the structures, one would expect the performance to depend on the complexity of the molecules.
We, however, find that the performance of the models does not correlate with complexity indicators but rather trivially with the size of the compounds (see \Cref{fig:performance_per_molecular_size}).
This indicates that the models might not (yet)  be able to reason about the structures of the molecules (in the way one might expect) but rather rely on retrieval of fragments of the training data.
Those observations mirror recent findings that \glspl{llm} struggle to \enquote{reverse} facts they have seen in training (i.e., generalize from \enquote{A has feature B} to \enquote{B is a feature of A}).\autocite{berglund2023reversal, zhu2023physics, allen2023physics, golovneva2024reverse}

@@

\begin{figure}[!h]
    \centering
    \includegraphics{figures/confidence_vs_performance_overall.pdf}
    \caption{\textbf{Relationship between confidence of the model in the answer and correctness.} For this analysis, we used verbalized confidence estimates from the model. To obtain those estimates, we prompted the models to return a confidence score on an ordinal scale. We then plot the correctness of the answers (which is calculated as the mean value of all answers being either completely correct (1) or completely incorrect (0)) against the confidence score.
        The stripplots show the individual data points, a strong density indicating a higher number of points. The lines show the mean and the errorbars show the standard error of the mean. The figure shows that the reliability of the confidence estimates varies widely across models. While the ones for GPT-4 and Claude 2 are not particularly reliable, the ones for Claude 3 follow the expected trend.}
    \label{fig:confidence_vs_performance}
    \script{joint_analysis_confidence_performance.py}
\end{figure}

@@

\section{Conclusions}
On one hand, our findings underline the impressive capabilities of \glspl{llm} in the chemical sciences: Leading models outperform domain experts in certain chemistry questions on many topics. 
On the other hand, there are still striking limitations. 
For very relevant topics the answers models provide are wrong. 
On top of that, many models are not able to reliably estimate their own limitations.
Yet, the success of the models in our evaluations perhaps also reveals more about the limitations of the exams we use to evaluate models---and chemists---than about the models themselves.
For instance, while models perform well on many textbook questions, they struggle with questions that require some more reasoning. 
Given that the models outperform the average human in our study, it is clear that we need to rethink the way we teach and examine chemistry.
Critical reasoning is increasingly important and rote solving of problems or memorization of facts is clearly a domain in which \glspl{llm} will continue to outperform humans.

Our findings also highlight the nuanced trade-off between breadth and depth of evaluation frameworks. 
The analysis of model performance on different topics shows that the performance of models varies widely across subfields they are tested on. 
But even within a topic, the performance of models can vary widely depending on the type of question and the reasoning required to answer it.

The current evaluation frameworks for \glspl{llm} are mostly designed to measure the performance of the models on specific property prediction tasks. 
They cannot be used to evaluate reasoning or systems built for scientific applications. 
Thus, we had only little understanding of the capabilities of \glspl{llm} in the chemical sciences.
Our work shows that carefully curated benchmarks can provide a more nuanced understanding of the capabilities of \glspl{llm} in the chemical sciences.
Importantly, our findings also illustrate that more focus is required in the development of better human-model interaction frameworks given that models are not able to estimate their own limitations.

While our findings indicate many areas for further improvement of \gls{llm}-based systems, it is also important to realize that clearly defined metrics have been the key for the progress of many fields of \gls{ml} such computer vision. 
We thus believe that although current systems are far from reasoning like a chemist, our \chembench framework will be a stepping stone for the development of systems that might come closer to this goal.

@@

\subsection{Curation workflow}\label{sec:curation}
For our dataset we curated questions from existing exams, but also programmatically created new questions.
Questions were added via Pull Requests on our GitHub repository and only merged into the corpus after passing manual review (\Cref{fig:curation_workflow}).

@@

\paragraph{Parsing}
Our parsing workflow is multistep and primarily based on regular expressions.
In the case of instruction-tuned models, we first attempt to identify the \texttt{[ANSWER][\textbackslash ANSWER]} environment we prompt the model to report the answer in.
In the case of completion models, this step is skipped. From there, we attempt to extract the relevant enumeration letters (for multiple-choice questions) or numbers.
In the case of numbers, our regular expression was engineered to be able to deal with various forms of scientific notation.
As initial tests indicated that models sometimes return integers in the form of words, e.g. \enquote{one} instead of \enquote{1}, we also implemented a word-to-number conversion using regular expressions.
In case these hard-coded parsing steps fail, we fall back to using a \gls{llm}, e.g. Claude 2, to parse the completion.
%Manual verification indicates that this \gls{llm}-based parsing represent not a relevant error source and has been performed correctly in all cases we checked (randomly sampled 4 questions per topic).

@@

M.A.\ expresses gratitude to the European Research Council (ERC) for evaluating the project with the reference number 101106377 titled \enquote{CLARIFIER} and accepting it for funding under the HORIZON TMA MSCA Postdoctoral Fellowships - European Fellowships. 
Furthermore, M.A.\ acknowledges the funding provided by UK Research and Innovation (UKRI) under the UK government’s Horizon Europe funding guarantee (Grant Reference: EP/Y023447/1; Organization Reference: 101106377).

M.R.\ and U.S.S.\ thanks the \enquote{Deutsche Forschungsgemeinschaft} for funding under the regime of the priority programme SPP 2363 \enquote{Utilization and Development of Machine Learning for Molecular Applications – Molecular Machine Learning} (SCHU 1229/63-1; project number 497115849).
